---
title: "LAGOS Spatial Analysis"
author: "Bri Rick"
date: "2/24/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

library(tidyverse) # Tidy packages
library(sf) #Spatial package that can read and create shapefiles 
library(mapview) #Interactive maps
library(LAGOSNE) #Lots and lots of clean lake data
library(USAboundaries) #USA states and counties
mapviewOptions(fgb = FALSE)
```


# LAGOS Analysis


## Loading in data


### First download and then specifically grab the locus (or site lat longs)

```{r data-read}
# #Lagos download script
#LAGOSNE::lagosne_get(dest_folder = LAGOSNE:::lagos_path())


#Load in lagos
lagos <- lagosne_load()

#Grab the lake centroid info
lake_centers <- lagos$locus

```



### Convert to spatial data
```{r}
#Look at the column names
#names(lake_centers)

#Look at the structure
#str(lake_centers)

#View the full dataset
#View(lake_centers %>% slice(1:100))

spatial_lakes <- st_as_sf(lake_centers,coords=c('nhd_long','nhd_lat'),
                          crs=4326) %>%
  st_transform(2163)

#Subset for plotting
subset_spatial <- spatial_lakes %>%
  slice(1:100) 

subset_baser <- spatial_lakes[1:100,]

#Dynamic mapviewer
mapview(subset_spatial, map.types = c("Esri.WorldImagery", "OpenStreetMap.DE"))

```


### Subset to only Minnesota

```{r}
states <- us_states()

#Plot all the states to check if they loaded
#mapview(states)
minnesota <- states %>%
  filter(name == 'Minnesota') %>%
  st_transform(2163)

#Subset lakes based on spatial position
minnesota_lakes <- spatial_lakes[minnesota,] %>% 
  mutate(state = "Minnesota")

#Plotting the first 1000 lakes
minnesota_lakes %>%
  arrange(-lake_area_ha) %>%
    slice(1:1000) %>%
  mapview(.,zcol = 'lake_area_ha', map.types = c("Esri.WorldImagery", "OpenStreetMap.DE"))
```



# In-Class work


## 1) Show a map outline of Iowa and Illinois (similar to Minnesota map upstream)

```{r}
#Subset states to just include Iowa and Illinois
IowaIllinois <- states %>% 
  filter(name %in% c("Iowa", "Illinois")) %>% 
  st_transform(2163)

#View outlines of Iowa and Illinois on the map
mapview(IowaIllinois, map.types = c("Esri.WorldImagery", "OpenStreetMap.DE"))
```



## 2) Subset LAGOS data to these sites, how many sites are in Illinois and Iowa combined? How does this compare to Minnesota?

There are 29,038 lakes located within Minnesota, whereas there are 16,466 lakes located in both Illinois and Iowa. Minnesota has almost 13,000 more lakes than the two states combined. 

```{r}
#Subset the lakes dataset by those located in Iowa and Illinois
IAIL_lakes <- spatial_lakes[IowaIllinois,]

#length(unique(IAIL_lakes$lagoslakeid))
```


## 3) What is the distribution of lake size in Iowa vs. Minnesota?

Minnesota has a much larger distribution of lakes than Iowa, meaning there are many more lakes, and more larger lakes than are present in Iowa. Otherwise they have a similar distribution with a distribution skewed toward small lakes.

```{r}
iowa <- states %>% 
  filter(name == "Iowa") %>% 
  st_transform(2163)

#Spatially subset lakes located in Iowa
iowa_lakes <- spatial_lakes[iowa,] %>% 
  mutate(state = "Iowa")

#Bind together MN and IA lake datasets
MNIAlakes <- rbind(minnesota_lakes, iowa_lakes)

#Plot a histogram of lake areas for both MN and IA
ggplot(MNIAlakes) +
  geom_histogram(aes(x = lake_area_ha, fill = state), alpha = 0.8, bins = 50, color = "black", show.legend = FALSE) +
  scale_x_log10() +
  theme_bw()+
  facet_grid(~state) +
  xlab("Lake area (Ha)")+
  ylab("Number of lakes")
```


## 4) Make an interactive plot of lakes in Iowa and Illinois and color them by lake area in hectares

```{r}
#Subset to view only 1000 lakes to make them easier to see
subset_IAIL <- IAIL_lakes %>%
  arrange(-lake_area_ha) %>%
  slice(1:1000) 

#Visualize subset of lakes on map
mapview(subset_IAIL, zcol = "lake_area_ha", map.types = c("Esri.WorldImagery", "OpenStreetMap.DE"))
```


## 5) What other data sources might we use to understand how reservoirs and natural lakes vary in size in these three states? 

We could map the area of the lakes using satellite imagery, which would also allow us to look at changes in area over time. 
